---
title: "Assignment 1: Air Transport Since 2013, Winners and Losers "
author: "Christina Moss and Hao Zhu"
date: "`r Sys.Date()`" 
output: 
  html_document:
    toc: true 
    theme: sandstone
    toc_float: true
    toc_depth: 2 
    code_folding: hide
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(dplyr)
library(ggplot2)
library(knitr)
library(kableExtra)
library(ggtext)
library(readr)
library(tidyr)
library(ggplot2)
```

## Exploring Financial Trends Across Airports with Shared Traits 

In this analysis we will examine financial and budgeting patterns at the Savannah/Hilton Head International Airport and the Charleston International Airport. The cities of Savannah, GA and Charleston, SC are both attractive tourist destinations in the coastal south with relatively similar population sizes. 

### Data Preparation 
```{r}
dat_2021 <- read_csv("Report127_2021.csv",
                     skip = 3,
                     na = "",
                     show_col_types = FALSE)
names(dat_2021) <- trimws(names(dat_2021))

dat_2022 <- read_csv("Report127_2022.csv",
                     skip = 3,
                     na = "",
                     show_col_types = FALSE)
names(dat_2022) <- trimws(names(dat_2022))

dat_2023 <- read_csv("Report127_2023.csv",
                     skip = 3,
                     na = "",
                     show_col_types = FALSE)
names(dat_2023) <- trimws(names(dat_2023))

```

```{r}
charleston_sav_2123 <- rbind(
  dat_2021 %>% 
    filter(`Airport Name` %in% c("CHARLESTON INTL", "SAVANNAH INTERNATIONAL")),
  dat_2022 %>% 
    filter(`Airport Name` %in% c("CHARLESTON INTL", "SAVANNAH INTERNATIONAL")),
  dat_2023 %>% 
    filter(`Airport Name` %in% c("CHARLESTON INTL", "SAVANNAH INTERNATIONAL")))

charleston_sav_2123 <- charleston_sav_2123 %>%
  mutate(
    Year = as.numeric(substr(FYE, nchar(FYE)-3, nchar(FYE)))  # Extract last 4 characters for year
  )

# Verify the result
print(charleston_sav_2123 %>% 
  select(`Airport Name`, FYE, Year) %>%
  arrange(Year, FYE))

```

### Exploring Revenue Breakdown Over Time 
```{r}

```

```{r}

```


### Exploring Cost Breakdown Over Time 
