---
title: "Assignment 1: Air Transport Since 2013, Winners and Losers "
author: "Christina Moss and Hao Zhu"
date: "`r Sys.Date()`" 
output: 
  html_document:
    toc: true 
    theme: sandstone
    toc_float: true
    toc_depth: 2 
    code_folding: hide
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(dplyr)
library(ggplot2)
library(knitr)
library(kableExtra)
library(ggtext)
```

# Part One: Analyzing Winning and Losing Airports (2013-2023)

This analysis examines changes in airport operations of major large and
small hubs over the past decade (2013-2023). Using data from the FAA's
Terminal Area Forecast, we define winners as airports that experienced
the largest percentage increases in commuter enplanements and,
conversly, losing airports as those which experienced the greatest
decline. Commuter enplanements is our key metric for measuring success
because it signifies an increase (or decrease) in connecting traffic,
which is an important indicator for defining an airport as a hub. More
connections means more total operations, signifying an airport with a
steady flow of traffic rather than the rush hour and lulls dynamic of
origin and destination airports.

We focus our analysis on percentage change because we wanted to
particularly highlight those airports experiencing the most growth (or
greatest decline). Identifying which airports have experienced the most
growth over the last decade can play a role in highlighting where
airlines should prioritize investment and expansion.

```{r data_load, include=FALSE}
dat_large <- read.delim("TAFDetailed.txt", sep="\t", header=TRUE)
dat_small <- read.delim("TAFDetailed_small.txt", sep="\t", header=TRUE)
```

## Defining Winners and Losers - Data Exploration

Here we identify large and small airport hubs with a commuter
enplanement number of greater than 100,000 passengers.

```{r data_explore, warning=FALSE, message=FALSE}

hub_large <- dat_large %>%
    filter(SYSYEAR == 2013,  COMMUTER > 100000)

hub_small <- dat_small %>%
  filter(SYSYEAR == 2023, COMMUTER > 100000)

large_change <- dat_large[dat_large$LOC_ID %in% hub_large$LOC_ID, ]
small_change <- dat_small[dat_small$LOC_ID %in% hub_small$LOC_ID, ]

ggplot(large_change, aes(x=SYSYEAR, y=COMMUTER, color=LOC_ID)) +
  geom_line() +
  theme_minimal() +
  labs(title="Airport Traffic Over Time (Large Hubs)",
       x="Year",
       y="Commuter Traffic",
       color="Airport") +
  scale_x_continuous(breaks=2013:2023)

ggplot(small_change, aes(x=SYSYEAR, y=COMMUTER, color=LOC_ID)) +
  geom_line() +
  theme_minimal() +
  labs(title="Airport Traffic Over Time (Small Hubs)",
       x="Year",
       y="Commuter Traffic",
       color="Airport") +
  scale_x_continuous(breaks=2013:2023)
```

The charts above highlight that Chicago and Boise are both the largest
hubs in their respective size categories. However, in order to identify
winners and losers, we are more interested in the change in commuter
enplanements for hub airports from 2013 to 2023.

```{r hub, fig.height=6, fig.width=8}
#planning to chart hubs with the largest percentage change from 2013 to 2023 to determine winners & losers 
large_change <- large_change %>%
  group_by(LOC_ID) %>%
  summarise(
    start_value = first(COMMUTER),
    end_value = last(COMMUTER),
    tot_change = end_value - start_value,
    pct_change = ((end_value - start_value) / start_value) * 100
  )

small_change <- small_change %>%
  group_by(LOC_ID) %>%
  summarise(
    start_value = first(COMMUTER),
    end_value = last(COMMUTER),
    tot_change = end_value - start_value,
    pct_change = ((end_value - start_value) / start_value) * 100
  )

# Horizontal Bar Chart - Percentage Change - Small Hubs
ggplot(small_change, aes(x = reorder(LOC_ID, pct_change), y = pct_change, fill = pct_change)) +
  geom_col() +
  coord_flip() +  
  labs(title = "Percent Change in Commuter Enplanements (Small Hubs)",
       x = "Airport",
       y = "Percent Change (%)") +
  scale_fill_gradient(low = "#FC6189", high = "#3BA4BF") +
  theme_minimal()
```

```{r hub2}
# Horizontal Bar Chart - Percent Change - Large Hubs
ggplot(large_change, aes(x = reorder(LOC_ID, tot_change), y = pct_change, fill = pct_change)) +
  geom_col() +
  coord_flip() +  
  labs(title = "Percent Change in Commuter Enplanements (Large Hubs)",
       x = "Airport",
       y = "Total Change (Passengers)") +
  scale_fill_gradient(low = "#FC6189", high = "#3BA4BF") +
  theme_minimal()

```

### Winners and Losers

-   Small hub winners (greatest positive percent change): FAI - SGF
-   Small hub losers : SBA - CID
-   Large hub winners (greatest positive percent change): LGA -JFK
-   Large hub losers: ORD - ATL

## Part Two: Airport Delay Patterns (2013-2023)

This analysis examines the operational efficiency of previously
identified winner and loser airports by comparing their delay metrics
between January 2013 and January 2023. Using data from the Office of
Airline Information, Bureau of Transportation Statistics (BTS), we
analyze both arrival and departure delays for each airport in their
roles as origin and destination points. Delay metrics serve as our key
indicators for measuring operational efficiency, as they reflect an
airport's ability to manage traffic flow and maintain service quality
regardless of their hub status. By examining these metrics across
different operational roles, we can better understand how changes in hub
status affect an airport's ability to handle traffic efficiently.

The focus on delay changes reveals an interesting contrast between hub
growth and operational efficiency. Among small hub winners, FAI and SGF
showed relatively modest delay increases despite their significant
growth in commuter enplanements - FAI actually improved its destination
arrival performance (-6.3 minutes), while SGF saw minimal increases
(0.2-3.0 minutes). Small hub losers displayed mixed performance: SBA
experienced the largest delay increases (8.1-12.2 minutes) while CID
showed improvement in destination metrics (-1.4 to -2.5 minutes). In the
large hub category, winners LGA and JFK faced operational challenges
with their growth, showing notable delay increases (JFK: 4.2-10.3
minutes), while large hub losers demonstrated divergent patterns - ATL
experienced significant delay increases (4.6-11.3 minutes) while ORD
maintained relatively stable operations with some improvements in origin
delays (-1.2 to -1.6 minutes). Our findings suggest that growth in hub
status doesn't necessarily correlate with decreased operational
efficiency.

```{r data_load2, include=FALSE}
dat_2013 <- read.csv("Performance_2013_Jan/Performance_2013_Jan.csv")
dat_2023 <- read.csv("Performance_2023_Jan/Performance_2023_Jan.csv")
```

```{r data_cleaning, include=FALSE}
selected_airports <- c("FAI", "SGF", "SBA", "CID", "LGA", "JFK", "ORD", "ATL")
performance_2013 <- dat_2013 %>%
  filter(Origin %in% selected_airports | Dest %in% selected_airports) %>%
  select(c("Year", "Month", "FlightDate", "Origin", "Dest", 
                     "DepDelay", "DepDelayMinutes", "DepDel15", 
                     "ArrDelay", "ArrDelayMinutes", "ArrDel15"))

performance_2023 <- dat_2023 %>%
  filter(Origin %in% selected_airports | Dest %in% selected_airports)%>%
  select(c("Year", "Month", "FlightDate", "Origin", "Dest", 
                     "DepDelay", "DepDelayMinutes", "DepDel15", 
                     "ArrDelay", "ArrDelayMinutes", "ArrDel15"))
```

```{r performance, warning=FALSE, message=FALSE}
# Combine datasets
combined_data <- bind_rows(
  performance_2013 %>% mutate(dataset = "2013"),
  performance_2023 %>% mutate(dataset = "2023")
)


airport_delays <- combined_data %>%
  # Create separate rows for origin and destination analysis
  pivot_longer(
    cols = c(Origin, Dest),
    names_to = "airport_role",
    values_to = "airport"
  ) %>%
  filter(airport %in% selected_airports) %>%
  group_by(dataset, airport, airport_role) %>%
  summarize(
    avg_dep_delay = mean(DepDelay, na.rm = TRUE),
    avg_arr_delay = mean(ArrDelay, na.rm = TRUE),
    n_flights = n()
  ) %>%
  ungroup()

# Create long format for plotting
delays_long <- airport_delays %>%
  pivot_longer(
    cols = c(avg_dep_delay, avg_arr_delay),
    names_to = "delay_type",
    values_to = "delay_minutes"
  )

# Visualization
ggplot(delays_long, aes(x = factor(airport, levels = selected_airports), 
                       y = delay_minutes, 
                       fill = dataset)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  facet_grid(airport_role ~ delay_type,
             labeller = labeller(
               delay_type = c(
                 "avg_dep_delay" = "Departure Delay",
                 "avg_arr_delay" = "Arrival Delay"),
               airport_role = c(
                 "Origin" = "As Origin Airport",
                 "Dest" = "As Destination Airport"))) +
  theme_minimal() +
  labs(
    title = "Average Flight Delays for Selected Airports: **<span style='color:#db7996;'>2013</span>** vs **<span style='color:#6a9fb8;'>2023</span>**",
    subtitle = "Analysis of delays when airport serves as origin vs destination",
    x = "Airport",
    y = "Average Delay (minutes)",
    fill = "Year"
  ) +
  scale_fill_manual(values = c("2013" = "#db7996", "2023" = "#6a9fb8")) +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 1),
    axis.title = element_text(size = 12),
    plot.title = element_markdown(size = 14, face = "bold"),
    legend.position = "none",
    panel.grid.minor = element_blank(),
    strip.text = element_text(size = 12, face = "bold")
  )
```

```{r summary, warning=FALSE, message=FALSE}
# Create summary statistics
summary_stats <- airport_delays %>%
  group_by(dataset, airport) %>%
  summarize(
    avg_dep_delay_as_origin = round(mean(avg_dep_delay[airport_role == "Origin"], na.rm = TRUE), 1),
    avg_arr_delay_as_origin = round(mean(avg_arr_delay[airport_role == "Origin"], na.rm = TRUE), 1),
    avg_dep_delay_as_dest = round(mean(avg_dep_delay[airport_role == "Dest"], na.rm = TRUE), 1),
    avg_arr_delay_as_dest = round(mean(avg_arr_delay[airport_role == "Dest"], na.rm = TRUE), 1),
    total_flights = sum(n_flights)
  ) %>%
  arrange(airport, dataset)


# Calculate year-over-year changes
delay_changes <- summary_stats %>%
  select(-total_flights) %>%
  pivot_wider(
    names_from = dataset,
    values_from = c(avg_dep_delay_as_origin, avg_arr_delay_as_origin,
                   avg_dep_delay_as_dest, avg_arr_delay_as_dest)
  ) %>%
  mutate(
    dep_delay_change_origin = avg_dep_delay_as_origin_2023 - avg_dep_delay_as_origin_2013,
    arr_delay_change_origin = avg_arr_delay_as_origin_2023 - avg_arr_delay_as_origin_2013,
    dep_delay_change_dest = avg_dep_delay_as_dest_2023 - avg_dep_delay_as_dest_2013,
    arr_delay_change_dest = avg_arr_delay_as_dest_2023 - avg_arr_delay_as_dest_2013
  )

```

```{r table, warning=FALSE, message=FALSE}
delay_table <- delay_changes %>%
  select(airport,
         dep_delay_change_origin,
         arr_delay_change_origin,
         dep_delay_change_dest,
         arr_delay_change_dest) %>%
  mutate(across(where(is.numeric), round, 1)) %>%
  rename(
    "Airport" = airport,
    "Departure Delay\n(as Origin)" = dep_delay_change_origin,
    "Arrival Delay\n(as Origin)" = arr_delay_change_origin,
    "Departure Delay\n(as Destination)" = dep_delay_change_dest,
    "Arrival Delay\n(as Destination)" = arr_delay_change_dest
  )

# Create a formatted table
kable(delay_table, format = "html", 
      caption = "Changes in Flight Delays (2023 vs 2013) in Minutes") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE, background = "#f0f0f0") %>%
  column_spec(1, bold = TRUE) %>%
  add_header_above(c(" " = 1, "As Origin" = 2, "As Destination" = 2)) %>%
  footnote(
    general = "Positive values indicate increased delays in 2023 compared to 2013",
    general_title = "Note: "
  )
```
