---
title: "Assignment 1: Air Transport Since 2013, Winners and Losers "
author: "Christina Moss and Hao Zhu"
date: "2025-02-03"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(dplyr)
library(ggplot2)
library(knitr)
library(kableExtra)
```

# Part One: Analyzing Winning and Losing Airports (2013-2023)

This analysis examines changes in airport operations of major large and
small hubs over the past decade (2013-2023). Using data from the FAA's
Terminal Area Forecast, we define winners as airports that experienced
the largest percentage increases in commuter enplanements and,
conversly, losing airports as those which experienced the greatest
decline. Commuter enplanements is our key metric for measuring success
because it signifies an increase (or decrease) in connecting traffic,
which is an important indicator for defining an airport as a hub. More
connections means more total operations, signifying an airport with a
steady flow of traffic rather than the rush hour and lulls dynamic of
origin and destination airports.

We focus our analysis on percentage change because we wanted to
particularly highlight those airports experiencing the most growth (or
greatest decline). Identifying which airports have experienced the most
growth over the last decade can play a role in highlighting where
airlines should prioritize investment and expansion.

```{r data_load, include=FALSE}
dat_large <- read.delim("TAFDetailed.txt", sep="\t", header=TRUE)
dat_small <- read.delim("TAFDetailed_small.txt", sep="\t", header=TRUE)
```

## Defining Winners and Losers - Data Exploration

Here we identify large and small airport hubs with a commuter
enplanement number of greater than 100,000 passengers.

```{r data_explore, warning=FALSE, message=FALSE}

hub_large <- dat_large %>%
    filter(SYSYEAR == 2013,  COMMUTER > 100000)

hub_small <- dat_small %>%
  filter(SYSYEAR == 2023, COMMUTER > 100000)

large_change <- dat_large[dat_large$LOC_ID %in% hub_large$LOC_ID, ]
small_change <- dat_small[dat_small$LOC_ID %in% hub_small$LOC_ID, ]

ggplot(large_change, aes(x=SYSYEAR, y=COMMUTER, color=LOC_ID)) +
  geom_line() +
  theme_minimal() +
  labs(title="Airport Traffic Over Time (Large Hubs)",
       x="Year",
       y="Commuter Traffic",
       color="Airport") +
  scale_x_continuous(breaks=2013:2023)

ggplot(small_change, aes(x=SYSYEAR, y=COMMUTER, color=LOC_ID)) +
  geom_line() +
  theme_minimal() +
  labs(title="Airport Traffic Over Time (Small Hubs)",
       x="Year",
       y="Commuter Traffic",
       color="Airport") +
  scale_x_continuous(breaks=2013:2023)
```

The charts above highlight that Chicago and Boise are both the largest
hubs in their respective size categories. However, in order to identify
winners and losers, we are more interested in the change in commuter
enplanements for hub airports from 2013 to 2023.

```{r hub}
#planning to chart hubs with the largest percentage change from 2013 to 2023 to determine winners & losers 
large_change <- large_change %>%
  group_by(LOC_ID) %>%
  summarise(
    start_value = first(COMMUTER),
    end_value = last(COMMUTER),
    tot_change = end_value - start_value,
    pct_change = ((end_value - start_value) / start_value) * 100
  )

small_change <- small_change %>%
  group_by(LOC_ID) %>%
  summarise(
    start_value = first(COMMUTER),
    end_value = last(COMMUTER),
    tot_change = end_value - start_value,
    pct_change = ((end_value - start_value) / start_value) * 100
  )

# Horizontal Bar Chart - Percentage Change - Small Hubs
ggplot(small_change, aes(x = reorder(LOC_ID, pct_change), y = pct_change, fill = pct_change)) +
  geom_col() +
  coord_flip() +  
  labs(title = "Percent Change in Commuter Enplanements (Small Hubs)",
       x = "Airport",
       y = "Percent Change (%)") +
  scale_fill_gradient(low = "red", high = "blue") +
  theme_minimal()

# Horizontal Bar Chart - Percent Change - Large Hubs
ggplot(large_change, aes(x = reorder(LOC_ID, tot_change), y = pct_change, fill = pct_change)) +
  geom_col() +
  coord_flip() +  
  labs(title = "Percent Change in Commuter Enplanements (Large Hubs)",
       x = "Airport",
       y = "Total Change (Passengers)") +
  scale_fill_gradient(low = "red", high = "blue") +
  theme_minimal()

```

Small hub winners (greatest positive percent change) - FAI - SGF 
Small hub losers - SBA - CID 
Large hub winners (greatest positive percent change) - LGA -JFK
Large hub losers - ORD - ATL



## Part Two: Delay and Departure Change in 2013 and 2023
```{r data_load2, include=FALSE}
dat_2013 <- read.csv("Performance_2013_Jan/Performance_2013_Jan.csv")
dat_2023 <- read.csv("Performance_2023_Jan/Performance_2023_Jan.csv")
```

```{r data_cleaning, include=FALSE}
selected_airports <- c("FAI", "SGF", "SBA", "CID", "LGA", "JFK", "ORD", "ATL")
performance_2013 <- dat_2013 %>%
  filter(Origin %in% selected_airports | Dest %in% selected_airports) %>%
  select(c("Year", "Month", "FlightDate", "Origin", "Dest", 
                     "DepDelay", "DepDelayMinutes", "DepDel15", 
                     "ArrDelay", "ArrDelayMinutes", "ArrDel15"))

performance_2023 <- dat_2023 %>%
  filter(Origin %in% selected_airports | Dest %in% selected_airports)%>%
  select(c("Year", "Month", "FlightDate", "Origin", "Dest", 
                     "DepDelay", "DepDelayMinutes", "DepDel15", 
                     "ArrDelay", "ArrDelayMinutes", "ArrDel15"))
```

```{r performance, warning=FALSE, message=FALSE}
# Combine datasets
combined_data <- bind_rows(
  performance_2013 %>% mutate(dataset = "2013"),
  performance_2023 %>% mutate(dataset = "2023")
)


airport_delays <- combined_data %>%
  # Create separate rows for origin and destination analysis
  pivot_longer(
    cols = c(Origin, Dest),
    names_to = "airport_role",
    values_to = "airport"
  ) %>%
  filter(airport %in% selected_airports) %>%
  group_by(dataset, airport, airport_role) %>%
  summarize(
    avg_dep_delay = mean(DepDelay, na.rm = TRUE),
    avg_arr_delay = mean(ArrDelay, na.rm = TRUE),
    n_flights = n()
  ) %>%
  ungroup()

# Create long format for plotting
delays_long <- airport_delays %>%
  pivot_longer(
    cols = c(avg_dep_delay, avg_arr_delay),
    names_to = "delay_type",
    values_to = "delay_minutes"
  )

# Visualization
ggplot(delays_long, aes(x = factor(airport, levels = selected_airports), 
                       y = delay_minutes, 
                       fill = dataset)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  facet_grid(airport_role ~ delay_type,
             labeller = labeller(
               delay_type = c(
                 "avg_dep_delay" = "Departure Delay",
                 "avg_arr_delay" = "Arrival Delay"),
               airport_role = c(
                 "Origin" = "As Origin Airport",
                 "Dest" = "As Destination Airport"))) +
  theme_minimal() +
  labs(
    title = "Average Flight Delays for Selected Airports: 2013 vs 2023",
    subtitle = "Analysis of delays when airport serves as origin vs destination",
    x = "Airport",
    y = "Average Delay (minutes)",
    fill = "Year"
  ) +
  scale_fill_manual(values = c("2013" = "#FF9999", "2023" = "#66C2A5")) +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 1),
    axis.title = element_text(size = 12),
    plot.title = element_text(size = 14, face = "bold"),
    legend.position = "bottom",
    panel.grid.minor = element_blank(),
    strip.text = element_text(size = 12, face = "bold")
  )
```


```{r summary, warning=FALSE, message=FALSE}
# Create summary statistics
summary_stats <- airport_delays %>%
  group_by(dataset, airport) %>%
  summarize(
    avg_dep_delay_as_origin = round(mean(avg_dep_delay[airport_role == "Origin"], na.rm = TRUE), 1),
    avg_arr_delay_as_origin = round(mean(avg_arr_delay[airport_role == "Origin"], na.rm = TRUE), 1),
    avg_dep_delay_as_dest = round(mean(avg_dep_delay[airport_role == "Dest"], na.rm = TRUE), 1),
    avg_arr_delay_as_dest = round(mean(avg_arr_delay[airport_role == "Dest"], na.rm = TRUE), 1),
    total_flights = sum(n_flights)
  ) %>%
  arrange(airport, dataset)

print(summary_stats)

# Calculate year-over-year changes
delay_changes <- summary_stats %>%
  select(-total_flights) %>%
  pivot_wider(
    names_from = dataset,
    values_from = c(avg_dep_delay_as_origin, avg_arr_delay_as_origin,
                   avg_dep_delay_as_dest, avg_arr_delay_as_dest)
  ) %>%
  mutate(
    dep_delay_change_origin = avg_dep_delay_as_origin_2023 - avg_dep_delay_as_origin_2013,
    arr_delay_change_origin = avg_arr_delay_as_origin_2023 - avg_arr_delay_as_origin_2013,
    dep_delay_change_dest = avg_dep_delay_as_dest_2023 - avg_dep_delay_as_dest_2013,
    arr_delay_change_dest = avg_arr_delay_as_dest_2023 - avg_arr_delay_as_dest_2013
  )

print(delay_changes)
```
```{r table, warning=FALSE, message=FALSE}
# Prepare the data for display
delay_table <- delay_changes %>%
  select(airport,
         dep_delay_change_origin,
         arr_delay_change_origin,
         dep_delay_change_dest,
         arr_delay_change_dest) %>%
  mutate(across(where(is.numeric), round, 1)) %>%
  rename(
    "Airport" = airport,
    "Departure Delay\n(as Origin)" = dep_delay_change_origin,
    "Arrival Delay\n(as Origin)" = arr_delay_change_origin,
    "Departure Delay\n(as Destination)" = dep_delay_change_dest,
    "Arrival Delay\n(as Destination)" = arr_delay_change_dest
  )

# Create a formatted table
kable(delay_table, format = "html", 
      caption = "Changes in Flight Delays (2023 vs 2013) in Minutes") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE, background = "#f0f0f0") %>%
  column_spec(1, bold = TRUE) %>%
  add_header_above(c(" " = 1, "As Origin" = 2, "As Destination" = 2)) %>%
  footnote(
    general = "Positive values indicate increased delays in 2023 compared to 2013",
    general_title = "Note: "
  )
```

